---
title: "Weather Event Damage Analysis Using NOAA Storm Database (1950~2011)"
author: "Junyoung Kim"
date: "Feb 11, 2018"
output: html_document
---

<br><br>
<br><br>

## 1. Synopsis

This study is to clarify **what kind of weather events happened in the U.S. have caused the biggest population health damages and economic losses**. Particularly, it answers to the two major questions below.

  1. Across the United States, which types of events (as indicated in the **EVTYPE** variable) are most harmful with respect to population health?
  2. Across the United States, which types of events have the greatest economic consequences?

To achieve this goal, this study examines the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm data. The NOAA data used in this study contains major and minor weather events between 1950 and 2011 and their direct and folowing effects on public health (fatalities and injuries) and on economy (property damages and crop damages). Although the [NOAA data instruction document](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) classifies the weather events into 48 different categories, the data itself has over 900 different categories including misspelled and duplicated ones, so this study reorganizes categories in order to make them a good fit to the guidelines proposed in the instruction document, and then summarizes accumulated casualties and economic losses of each category. As the result of the analyses, **Tornadoes** turned out to be the biggest cause of **health damages**, and **Floods** recorded the largest **total economic losses**. 

<br><br>
<br><br>

## 2. Data Processing

### 2.1. Downloading Data

The following R script downloads and reads the NOAA storm data. Before downloading and reading, the script checks whether the target file already exists, and sets locale to English in advance of reading data to prevent possible misinterpretation which can be caused by different language settings.

```{r, 2.1. Downloading Data}

Sys.setlocale("LC_ALL", "English") # set LOCALE
options(scipen=999) # turn scientific notation off

# set file URL and location
URL <-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
fileName <- "StormData.csv.bz2"
folder <- getwd()

# download and import
if(!file.exists(paste(folder, "/", fileName, sep="")))
        {download.file(URL, fileName)}

if(!exists("StormData")){StormData <- read.csv("StormData.csv.bz2", 
                                               stringsAsFactors = F)}
str(StormData)
dim(StormData)
```
<br><br>

### 2.2. Tidying Up Messy Event Types

Unfortunately, the raw NOAA data is not in the prime condition to analyze. The types of weather events, which should have been categorized in 48 kinds (see [NOAA data instruction document](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) p.6, Table 1), reaches 985 types including typos, different namings on the same categories, and events and summaries which are not listed in the instruction. To analyze the data, we need to correct typos and tidy up the messy event types. **The standard of recategorization follows the details in the instruction document.**  

```{r, 2.2. Data Processing: tidy up messy event types and correct typos}
StormData$EVTYPE <- toupper(StormData$EVTYPE) # capitalize all

# correct marks
a <- StormData$EVTYPE #duplicate first
a <- gsub("\\(.*\\)", "", a) # remove texts and numbers with parenthesis
a <- gsub("/", " ", a) # replace '/'s with spaces
a <- gsub("-", " ", a) # replace '-'s with spaces
a <- gsub("\\\\", " ", a) #replace '\\'s with spaces
a <- gsub("  ", " ", a) # remove double spaces
a <- gsub(";", " ", a) # remove semicolons
a <- gsub("\\.$", "", a) #remove dots at the end of the strings
a <- gsub("S$", "", a) # remove plural 'S's
a <- gsub("^ ", "", a) #remove blanks at the head of the strings
a <- gsub(" $", "", a) #remove blanks at the end of the strings

# tidy up labels and correct typos
# !!CAUTION!! DO NOT CHANGE THE ORDER BELOW
a[grep("AVA", a)] <- "AVALANCHE"
a[grep("BLIZ", a)] <- "BLIZZARD"
a[grep("^(?=.*FLOOD)(?=.*COAS)|(?=.*CSTL)|(?=.*BEACH)", 
                                               a, perl=TRUE)] <- "COASTAL FLOOD"
a[grep("EXTREME ",a)] <- "EXTREME COLD/WIND CHILL"
a[grep("DENSE FOG|^FOG$", a)] <- "DENSE FOG"
a[a=="FOG AND COLD TEMPERATURE"|a=="FREEZING FOG"|a=="ICE FOG"]<- "FREEZING FOG"
a[grep("SMOKE", a)] <- "DENSE SMOKE"
a[grep("DROUGHT", a)] <- "DROUGHT"
a[grep("DUST D", a)] <- "DUST DEVIL"
a[grep("G DUST|DUST S|DUSTS", a)] <- "DUST STORM"
a[grep("E HEAT", a)] <- "EXCESSIVE HEAT"
a[grep("^HEAT|D HEAT", a)] <- "HEAT"
a[grep("^(?=.*HEAVY)(?=.*SNOW)", a, perl=TRUE)] <- "HEAVY SNOW"
a[grep("^(?=.*HEAVY)(?=.*RAIN)|(?=.*PRECI)", a, perl=TRUE)] <- "HEAVY RAIN"
a[grep("SLEET", a)] <- "SLEET"
a[grep("FROST", a)] <- "FROST/FREEZE"
a[grep("^(?=.*FREEZ)(?!.*RAIN)(?!.*SNOW)(?!.*FOG)", 
                                                a, perl=TRUE)] <- "FROST/FREEZE"
a[grep("^(?=.*FREEZ)(?=.*RAIN)", a, perl=TRUE)] <- "WINTER WEATHER"
a[grep("^(?=.*TORN)(?!.*^WATER)", a, perl=TRUE)] <- "TORNADO"
a[a=="LANDSPOUT"] <- "TORNADO"
a[grep("^(?=.*THUN)|(?=.*ERSTO)", a, perl=TRUE)] <- "THUNDERSTORM WIND"
a[grep("^(?=.*TSTM)(?!.*^MARINE)", a, perl=TRUE)] <- "THUNDERSTORM WIND"
a[grep("(?=.*FUNN)(?!.*WATER)", a, perl=TRUE)] <- "FUNNEL CLOUD"
a[grep("^(?=.*HAIL)(?!.*MARINE)", a, perl=TRUE)] <- "HAIL"
a[grep("CURRENT", a)] <- "RIP CURRENT"
a[grep("SURF", a)] <- "HIGH SURF"
a[grep("HURR|TYPH", a)] <- "HURRICANE/TYPHOON"
a[grep("ICE STORM", a)] <- "ICE STORM"
a[grep("(?=.*FLOOD)(?=.*FLASH)", a, perl=TRUE)] <- "FLASH FLOOD"
a[grep("^(?=.*FLOOD)(?!.*FLASH)(?!.*COAST)(?!LAKES)", a, perl=TRUE)] <- "FLOOD"
a[grep("(?=.*URB)(?=.*SM)", a, perl=TRUE)] <- "FLOOD"
a[grep("LIGHTNING|LIGHTING", a)] <- "LIGHTNING"
a[a=="MARINE TSTM WIND"] <- "MARINE THUNDERSTORM WIND"
a[a=="STORM SURGE"|a=="STORM SURGE TIDE"] <- "STORM SURGE/TIDE"
a[grep("^(?=.*TROP)(?=.*STORM)", a, perl=TRUE)] <- "TROPICAL STORM"
a[grep("^(?=.*COAST)(?=.*STORM)", a, perl=TRUE)] <- "TROPICAL STORM"
a[grep("(?=.*VOL)(?=.*ASH)", a, perl=TRUE)] <- "VOLCANIC ASH"
a[grep("SPOUT", a)] <- "WATERSPOUT"
a[grep("FIRE", a)] <- "WILD FIRE"
a[grep("WINTER STORM|SNOWSTORM", a)] <- "WINTER STORM"
a[grep("(?=.*BURST)|(?=.*GUSTN)", a, perl=TRUE)] <- "THUNDERSTORM WIND"
a[grep("(?=.*RAIN)(?!.*LOW)(?!.*SNOW)", a, perl=TRUE)] <- "HEAVY RAIN"
a[grep("(?=.*COLD)(?!.*EXTR)|(?=.*CHIL)(?!.*EXTR)", 
                                             a, perl=TRUE)] <- "COLD/WIND CHILL"
a[grep("(?=.*WIND)(?=.*STORM)(?!.*MARIN)", a, perl=TRUE)] <- "THUNDERSTORM WIND"
a[a=="STORM FORCE WIND"] <- "THUNDERSTORM WIND"
a[a=="WHIRLWIND"] <- "TORNADO"
a[grep("^(?=.*HIGH WIND)(?!.*MARINE)", a, perl=TRUE)] <- "HIGH WIND"
a[grep("^(?=.*WIND)(?!.*MAR)(?!.*CHI)(?!.*HIG)(?!.*STO)(?!.*SNO)", 
                                                 a, perl=TRUE)] <- "STRONG WIND"
a[grep("(?=.*SNOW)(?=.*EXC)|(?=.*SNOW)(?=.*REC)", a, perl=TRUE)] <- "HEAVY SNOW"
a[grep("(?=.*SNOW)(?!.*HEA)(?!.*LAK)", a, perl=TRUE)] <- "WINTER WEATHER"
a[grep("ICE", a)] <- "WINTER WEATHER"
a[a=="WND"] <- "STRONG WIND"
a[grep("(?=.*WIN)(?=.*MIX)(?!.*STO)", a, perl=TRUE)] <- "WINTER WEATHER"
a[grep("HIGH TEMP", a)] <- "HIGH TEMPERATURE"
a[grep("WARM", a)] <- "WARM TEMPERATURE"
a[grep("WET", a)] <- "WET WEATHER"
a[grep("LOW RA", a)] <- "DRYNESS"
a[grep("DRY|DRI", a)] <- "DRYNESS"
a[grep("HOT", a)] <- "HIGH TEMPERATURE" 
a[grep("(?=.*COOL)|(?=.*LOW)(?!.*TIDE)", a, perl=TRUE)] <- "LOW TEMPERATURE"

#apply changes
StormData$EVTYPE <- a
rm(a)

```

<br><br>

### 2.3. Unifying Economic Damage Units

Because the units of property and crop damages in the data vary, we need to unify them to calculate total economic effects. According to the [NOAA data instruction document](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) (p. 12), To record all estimated costs within three digits, different units("PROPDMGEXP", "CROPDMGEXP") are given: for example, "K" for thousands, "M" for millions, "B" for billions. As we can see in the names of the unit columns, the numbers in the unit vectors mean exponents of 10. The following code chunk converts all different units into dollars, and calculates dollar-based damage costs of each event in the data by multiplying three-digit records with converted units.

```{r, 2.3. Unifying Economic Damage Units}
# Unify the units of the damage costs in the vectors
table(StormData$PROPDMGEXP) ; table(StormData$CROPDMGEXP) # check types of exps

# create lists of exps
# reverse vectors orders in order to prevent 0 -> 10^0 -> 10^1
units <- rev(names(table(StormData$PROPDMGEXP)))
exps <- rev(c(0, 0, 0, 1, 10^0, 10^1, 10^2, 10^3, 10^4, 10^5, 10^6, 10^7, 10^8, 
          10^9, 10^2, 10^2, 10^3, 10^6, 10^6))

# replace values
for(i in 1:19)
        {StormData$PROPDMGEXP[StormData$PROPDMGEXP == units[i]] <- exps[i]
         StormData$CROPDMGEXP[StormData$CROPDMGEXP == units[i]] <- exps[i]}

# deal with "k"s in CROPDMGEXP
StormData$CROPDMGEXP[StormData$CROPDMGEXP == "k"] <- 10^3 

# calculate economic damages
prop.damages <- StormData$PROPDMG * as.numeric(StormData$PROPDMGEXP)
crop.damages <- StormData$CROPDMG * as.numeric(StormData$CROPDMGEXP)

# make a tidy data
TidyData <- data.frame(as.factor(StormData$EVTYPE), # factorize event types 
                   StormData$FATALITIES, # population health #1
                   StormData$INJURIES,   # population health #2
                   prop.damages, # property damages
                   crop.damages) # crop damages
```

<br><br>
<br><br>

## 3. Results 

### 3.1. Answering Qustion #1: The biggest cause of casualties  

To answer the question, the codes below sum up the fatalities and injuries of each event type and create a plot of top 10 weather event types inducing health damages.

```{r, 3.1.}

# data for casualties
Casualty.Data <- data.frame(tapply(TidyData[,2], TidyData[,1], sum),
                            tapply(TidyData[,3], TidyData[,1], sum))
Casualty.Data$Total <- rowSums(Casualty.Data)
Casualty.Data <- Casualty.Data[order(-Casualty.Data$Total),]
colnames(Casualty.Data) <- c("Fatalities", "Injuries", "Total")

# top 10 causes
Top.Casualty <- Casualty.Data[1:10,]

# change plot margins
par(mar=c(10.1, 5.1, 4.1, 2.1))

# plot casualties
barplot(Top.Casualty$Total, names.arg=rownames(Top.Casualty), las=2, 
        cex=0.8, cex.axis=0.9,
        main="Top 10 Weather Events Causing Health Damages",
        col=c("brown1", rep("mistyrose2", 9)))
mtext(side = 2, "Total Casualty", line = 4)

Top.Casualty[1,3] # tornado
sum(Top.Casualty[2:10,3]) # sum of other top 10 events 
```

As we can see in the plot, Tornadoes recorded the largest number of total casualties(fatalities and injuries), therefore, **tornadoes are the most harmful with respect to population health.** Total casualties by tornadoes is about 97000, which exceeds the twice of the casualties caused by the other events in the top 10 lists(45208).
<br><br>

### 3.2. Answering Qustion #2: The biggest cause of economic losses  

Likewise, the code chunk below sums up the property and crop damages of each event type and create a plot of top 10 weather event types inducing economic damages.

```{r, 3.2.}
# change plot margins
par(mar=c(10.1, 5.1, 4.1, 2.1))

# data for economic losses
Economy.Data <- data.frame(tapply(TidyData[,4], TidyData[,1], sum),
                           tapply(TidyData[,5], TidyData[,1], sum))
Economy.Data$Total <- rowSums(Economy.Data)
Economy.Data <- Economy.Data[order(-Economy.Data$Total),]
colnames(Economy.Data) <- c("Property Damages", "Crop Damages", "Total")

# top 10 causes
Top.Economy <- Economy.Data[1:10,]

# plot economic losses
barplot(round(Top.Economy$Total / 10^6), 
        names.arg=rownames(Top.Economy), las=2, 
        cex=0.8, cex.axis=0.9,
        main="Top 10 Weather Events Causing Economic Losses",
        col=c("turquoise4", rep(rgb(157,209,203, maxColorValue = 255), 9)))
mtext(side = 2, "Total Loss  (Million U.S.$)", line = 4)

Economy.Data[1,3] # the biggest - FLOOD
Economy.Data[2,3] # the second biggest - HURRICANE/TYPHOON
Economy.Data[3,3] # the third biggest - TORNADO

# reset plot margins
par(mar=c(5.1, 4.1, 4.1, 2.1))
```

As we can see in the plot, **Floods caused over 150 billion dollar losses for the 60 years recorded($161 billion)**, and it is far beyond of the second (HURRICANE/TYPHOON, $90 Billion), and the third (TORNADO, $58 Billion).  

<br><br>
<br><br>
